---
title: "The Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Paolo Saracco"
date: "`r Sys.Date()`"
output:
        bookdown::pdf_document2:
                fig_caption: true
                toc: false
                number_sections: false
                keep_md: true
---

# Synopsis

Based on the `ToothGrowth` data from the R `datasets` package, we analyze
the effect of vitamin C on the length of odontoblasts in guinea pigs. We want 
to study whether the amount of vitamin C or the method it is delivered to 
guinea pigs affect the length of the odontoblasts. We conjecture that they do.

# Data 

The dataset contains the length of odontoblasts (cells responsible for tooth
growth) in 60 guinea pigs that received different doses of vitamin C by 
different methods. Each animal received one of three dose levels of
vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice 
or ascorbic acid (a form of vitamin C and coded as VC).

```{r data}
set.seed(0);
data("ToothGrowth");
```

### Packages

The following packages will be used to perform the analysis.

```{r packages, message = FALSE}
library(dplyr);
library(ggplot2);
```

# Exploratory analysis

We start by having a look at the size of our data set: 
`r paste("dim =", dim(ToothGrowth)[1], "x", dim(ToothGrowth)[2], sep = " ")`.
Hence we have, as expected, `r dim(ToothGrowth)[2]` observations for 
`r dim(ToothGrowth)[1]` subjects. Let us have a look at the head, the tail 
and the structure of our data

```{r datainspection, echo = F}
head(ToothGrowth, n = 3);
tail(ToothGrowth, n = 3);
str(ToothGrowth);
```

respectively. The dataset is already clean and we may provide a basic summary of the data:

```{r summary, echo = F}
summary(ToothGrowth)
```

By keeping in mind our working questions, we can have a quick look at 
potential correlations:

```{r pairplot, echo = F, fig.align = "center", out.height = "2in", fig.pos = "H"}
pairs(ToothGrowth)
```

Apparently, the dose level has a clear effect, while we cannot draw conclusions
about the delivery method, yet. The following additional boxplot, grouped by 
delivery method, confirms the effect of dosage and suggest a possible effect 
of delivery method at dose level 0.5 and 1:

```{r boxplot1, echo = F, fig.align = "center", out.height = "2in", fig.pos = "H"}
ToothGrowth <- ToothGrowth %>% mutate(dose = factor(dose), supp = factor(supp))
g <- ggplot(ToothGrowth, aes(x = dose, y = len, fill = dose))
g <- g + geom_boxplot() + facet_wrap(~ supp) + theme_bw()
g <- g + ggtitle("Odontoblast Length vs Dose level by Supplement type ") +
        labs(x = "Dose (mg/day)", y = "Length") +
        guides(fill = guide_legend(title = "Dose"))
g
```

# Hypothesis tests

In view of the relative small size of samples and under the reasonable 
hypothesis that the underlying data are iid Gaussian, we are going to perform 
two group T tests for equality of means. However, it is important to consider the 
multiple hypothesis testing issue here as we are going to perform 5
tests. We will implicitly apply the Bonferroni correction, so for a level 
\(\alpha = 0.05\) we will consider significant the p-values under \(0.01\).

### Compare tooth growth by dose level

Our first hypothesis is that a higher dose level of vitamin C leads to a 
higher length of odontoblasts. To support this hypothesis we compare level 
1 with level 0.5 and level 2 with level 1, the null hypothesis always being 
that there is no significant difference in the means and the alternative
being that a higher level of vitamin C entails a longer length of odontoblasts.

First we split the three groups according to the rule 

- group 1: dose level = 0.5 mg/day,
- group 2: dose level = 1 mg/day,
- group 3: dose level = 2 mg/day

and we compute their sample means:

```{r splitDose, echo = F}
grp1 <- ToothGrowth %>%
        filter(dose == 0.5) %>%
        select(len)

grp2 <- ToothGrowth %>%
        filter(dose == 1) %>%
        select(len)

grp3 <- ToothGrowth %>%
        filter(dose == 2) %>%
        select(len)

c("grp1" = mean(grp1$len), "grp2" = mean(grp2$len), "grp3" = mean(grp3$len))
```

Then we perform a T test to check whether there is a significant difference 
between the means at level \(\alpha = 0.05\). For the sake of space, 
we only report the p-value. The full output of the test can be found in the 
appendices. If we compare group 2 (dose 
level = 1 mg/day) with group 1 (dose level = 0.5 mg/day):

```{r firstTest}
t.test(grp2, grp1, alternative = "greater")$p.value
```

We can reject the null hypothesis that the two means coincide, since the 
p-value is much smaller than 0.01. Similarly, if we compare group 3 (dose 
level = 2 mg/day) with group 2 (dose level = 1 mg/day):

```{r secondTest}
t.test(grp3, grp2, alternative = "greater")$p.value
```

then we can reject the null hypothesis of equal means, since the p-value is much 
smaller than 0.01. There is no need, in this case, to test group 3 versus 
group 1.

We conclude that a higher level of vitamin C is related with a 
longer length of the odontoblasts at level 0.05.

### Compare tooth growth by delivery method

Our second hypothesis is that also the delivery method affects 
tooth growth. Namely, we suspect that subjects that received vitamin C at dose 
levels 0.5 and 1 mg/day through orange juice have longer odontoblasts. Instead, 
we believe that the supply method has no significant effect at a dosage of 
2 mg/day.

We separate the three groups by dosage, OJ versus VC, and we 
compute the respective sample means:

```{r suppsGrp, echo = F}
suppOJgrp1 <- ToothGrowth %>%
        filter(supp == "OJ", dose == "0.5") %>%
        select(len)

suppOJgrp2 <- ToothGrowth %>%
        filter(supp == "OJ", dose == "1") %>%
        select(len)

suppOJgrp3 <- ToothGrowth %>%
        filter(supp == "OJ", dose == "2") %>%
        select(len)

suppVCgrp1 <- ToothGrowth %>%
        filter(supp == "VC", dose == "0.5") %>%
        select(len)

suppVCgrp2 <- ToothGrowth %>%
        filter(supp == "VC", dose == "1") %>%
        select(len)

suppVCgrp3 <- ToothGrowth %>%
        filter(supp == "VC", dose == "2") %>%
        select(len)

c("OJgrp1" = mean(suppOJgrp1$len), "OJgrp2" = mean(suppOJgrp2$len), 
  "OJgrp3" = mean(suppOJgrp3$len), "VCgrp1" = mean(suppVCgrp1$len), 
  "VCgrp2" = mean(suppVCgrp1$len), "VCgrp3" = mean(suppVCgrp3$len))
```

Then we perform a T test to check whether there is a significant difference 
between the means at level \(\alpha = 0.05\). If we compare the OJ group with the VC group:

```{r thirdTest}
t.test(suppOJgrp1,suppVCgrp1,alternative = "greater")$p.value
t.test(suppOJgrp2,suppVCgrp2,alternative = "greater")$p.value
```

then we can reject the hypothesis of equal means at level 0.05 for 
the 0.5 and the 1 mg/day dose levels, while

```{r fourthTest}
t.test(suppOJgrp3,suppVCgrp3,alternative = "greater")$p.value
```

so we cannot reject the null hypothesis of equal means for the highest dosage.

We conclude that there is a relationship between the delivery method and 
the length of the odontoblasts at lower dosage (0.5 or 1 mg/day): subjects that 
received vitamin C via orange juice have, on average, longer odontoblasts than 
those who received it via ascorbic acid.

# Appendices

## Full outcomes of the T tests to check the effect of dosage.

### Group 2 (dose level = 1 mg/day) vs group 1 (dose level = 0.5 mg/day):

```{r firstTestout}
t.test(grp2, grp1, alternative = "greater")
```

### Group 3 (dose level = 2 mg/day) vs group 2 (dose level = 1 mg/day):

```{r secondTestout}
t.test(grp3, grp2, alternative = "greater")
```

## Full outcomes of the T tests to check the effect of delivery method by dosage

### Orange juice vs ascorbic acid at dosage 0.5

```{r thirdTestout1}
t.test(suppOJgrp1,suppVCgrp1,alternative = "greater")
```

### Orange juice vs ascorbic acid at dosage 1

```{r thirdTestout2}
t.test(suppOJgrp2,suppVCgrp2,alternative = "greater")
```

### Orange juice vs ascorbic acid at dosage 2

```{r fourthTestout}
t.test(suppOJgrp3,suppVCgrp3,alternative = "greater")
```

Readers interested in a literate version of this report, with explicit R code,
may refer to the 
[R markdown file](https://github.com/paolosaracco/DataScienceSpecialization/blob/main/06_StatisticalInference/CourseProject/BasicInferentialDataAnalysis.Rmd) 
on my GitHub repository. 

